A Pandemic-Scale Ancestral Recombination Graph for SARS-CoV-2

### 2025-11-07 latex/main.tex
Updates from Overleaf

\author[1,7,*,\orcidlink{0000-0002-6351-8873}]{Shing~H.~Zhan}
\author[1,6,\orcidlink{0000-0002-3536-6411}]{Yan~Wong}
\author[2,6,\orcidlink{0000-0001-6402-9396}]{Anastasia~Ignatieva}
\author[3]{Katherine~Eaton}
\author[1,\orcidlink{0000-0002-3547-7226}]{Isobel~Guthrie}
\author[1,\orcidlink{0000-0002-1982-6801}]{Benjamin~Jeffery}
\author[1,2,4,\orcidlink{0000-0002-0824-0047}]{Duncan~S.~Palmer}
\author[3,\orcidlink{0000-0002-1543-4501}]{Carmen~Lia~Murall}
\author[5,\orcidlink{0000-0003-3042-0818}]{Sarah~P.~Otto}
\author[1,**,\orcidlink{0000-0002-7894-5253}]{Jerome~Kelleher}

@@

\affil[1]{Big Data Institute, Li Ka Shing Centre for Health
Information and Discovery, University of Oxford, United Kingdom}
\affil[2]{Department of Statistics, University of Oxford, United Kingdom}
\affil[3]{National Microbiology Laboratory, Public Health Agency of Canada, Canada}
\affil[4]{The Pioneer Centre for SMARTbiomed, Big Data Institute, Li Ka Shing Centre for Health Information and Discovery, University of Oxford, United Kingdom}
\affil[5]{Department of Zoology and Biodiversity Research Centre, University of British Columbia, Vancouver BC Canada}

\affil[6]{Joint second author}
\affil[7]{Lead contact}

@@

\section*{DISCUSSION}
There are two primary contributions of this work. The first is an
ARG for 2.48 million SARS-CoV-2 genomes.
Through multiple layers of validation against results from the 
literature, we have demonstrated that this ARG synthesises
the evolutionary features of genetic inheritance, point mutations,
deletions and recombination in a single cohesive structure.
The ARG is freely available and easily accessible using the 
tskit software ecosystem, a rich and efficient platform for 
recombination-aware analyses widely used in population genomics.
This combination of a deeply validated pandemic-scale ARG with 
a mature and highly capable software platform forms a unique resource.
Previous landmark ARGs~\cite{Wohns2022,Anderson2023genes}
have helped spur the development of ARG-aware methods in human 
genomics~\cite{Mejia2025using,Grundler2025geographic,
Lehmann2025args,Terhorst2025accelerated}.
We hope that the ARG provided here will similarly
enable a range of recombination-aware analyses of SARS-CoV-2, and 
catalyse a wave of methodological developments integrating
recombination into pathogen genomics.
For instance, annotating the ARG with experimentally derived
fitness estimates for mutations can help elucidate selection
pressures on different recombinants;
analysing the genomic positions of mutations occurring on long
branches (which persist after accounting for recombination) can
illuminate the evolution of saltational variants.
We envisage the ARG as a long-term stable resource---building on the 
high-quality, freely available reassemblies of
the Viridian dataset~\cite{Hunt2024}---to facilitate the study 
of the pandemic phase of SARS-CoV-2 in an open and reproducible manner.
% NOTE: still missing the "What" bit suggested by Jotun, where we 
% list out some of the stuff that could be done. I'm not convinced
% this works, though.

The other primary contribution of this work is the method used to 
infer the ARG, sc2ts. The ability to infer an ARG over millions of whole 
genomes is a step change in scale over existing ARG 
inference methods for pathogen data\cite{Didelot2010inference,%
Vaughan2017inferring,%
Muller2020bayesian,%
Ignatieva2022,%
Muller2022bayesian}.
To ensure these innovations are well-founded, we have paid significant 
attention to methodological detail, carefully validating the ARG
(and consequently the method) against known results. 
Sc2ts is currently specialised to take advantage of unique features 
of the SARS-CoV-2 dataset, but the methods may be readily adapted
to other pathogens. The LS HMM is the most important element of 
sc2ts, efficiently finding parsimonious placements for samples 
among all possible recombinant solutions. The current 
parametrisation---emphasising interpretability---is simplistic,
and could readily be extended to account
for mutation rate variation along the genome~\cite{Haddox2025}
or transition-to-transversion rate bias~\cite{Bloom2023,Simmonds2024}.
Our analysis of deletions shows that valuable phylogenetic signal
is lost when they are ignored, and an important avenue for 
future work is to explore ways of incorporating them into the HMM.
Another methodological innovation is the use of 
the tsdate method~\cite{Wohns2022,Pope2025} to estimate internal node dates
in the ARG. While results are encouraging, 
dating of phylogenies is a complex task~\cite{Kumar2016advances},
methods for incorporating information from metadata of uncertain 
quality into large genealogies are in their 
early stages~\cite{sanderson2021chronumental}, and there is certainly
substantial room for improvement.
Sc2ts itself may benefit from these node dates, as 
much of the method's complexity lies in dealing with 
sample with incorrect sampling dates, and the ARG may provide 
a means of pre-identifying such ``time travellers''.

Further contributions of this work arise from 
the detailed study of the sc2ts ARG,
providing insights into the evolution of SARS-CoV-2
and the COVID-19 pandemic. 
First, we clarify the relationships among
recombinant Pango X lineages, finding examples of
clustering of named lineages under a single shared recombination event.
Second, we used the ARG to study breakpoint intervals 
for recombination events, adding a novel temporal dimension
and demonstrating a simple relationship with the 
size of the interval in base pairs derived from theory.
Thirdly, we demonstrate a clear relationship between
genetic diversity and the rate of detectable recombination,
with the best predictor of the number of recombination
events over time being the product of global case numbers
and lineage diversity.
Finally, and perhaps most fundamentally, through our 
conservative parameter choices and real-time handling of 
recombination events, we provide a stringent lower
bound on the amount of recombination that occurred during
the pandemic phase of SARS-CoV-2.
Together, these results underscore the utility of the ARG itself 
and suggest important directions for future research.

>>>>

\author[1,2,8,*,\orcidlink{0000-0002-6351-8873}]{Shing~H.~Zhan}
\author[1,7,\orcidlink{0000-0002-3536-6411}]{Yan~Wong}
\author[3,7,\orcidlink{0000-0001-6402-9396}]{Anastasia~Ignatieva}
\author[4]{Katherine~Eaton}
\author[1,\orcidlink{0000-0002-3547-7226}]{Isobel~Guthrie}
\author[1,\orcidlink{0000-0002-1982-6801}]{Benjamin~Jeffery}
\author[1,3,5,\orcidlink{0000-0002-0824-0047}]{Duncan~S.~Palmer}
\author[4,\orcidlink{0000-0002-1543-4501}]{Carmen~Lia~Murall}
\author[6,\orcidlink{0000-0003-3042-0818}]{Sarah~P.~Otto}
\author[1,**,\orcidlink{0000-0002-7894-5253}]{Jerome~Kelleher}

@@

\affil[1]{Big Data Institute, Li Ka Shing Centre for Health
Information and Discovery, University of Oxford, United Kingdom}
\affil[2]{Infectious Disease Epidemiology Unit (IDEU), Nuffield Department of Population Health, University of Oxford, Oxford, United Kingdom}
\affil[3]{Department of Statistics, University of Oxford, United Kingdom}
\affil[4]{National Microbiology Laboratory, Public Health Agency of Canada, Canada}
\affil[5]{The Pioneer Centre for SMARTbiomed, Big Data Institute, Li Ka Shing Centre for Health Information and Discovery, University of Oxford, United Kingdom}
\affil[6]{Department of Zoology and Biodiversity Research Centre, University of British Columbia, Vancouver BC Canada}

\affil[7]{Joint second author}
\affil[8]{Lead contact}

@@

\section*{DISCUSSION}
There are two primary contributions of this work. The first is an
ARG for 2.48 million SARS-CoV-2 genomes.
Through multiple layers of validation against results from the 
literature, we have demonstrated that this ARG synthesises
the evolutionary features of genetic inheritance, point mutations,
deletions and recombination in a single cohesive structure.
The ARG is freely available and easily accessible using the 
tskit software ecosystem, a rich and efficient platform for 
recombination-aware analyses widely used in population genomics.
This combination of a deeply validated pandemic-scale ARG with 
a mature and highly capable software platform forms a unique resource.
Previous landmark ARGs~\cite{Wohns2022,Anderson2023genes}
have helped spur the development of ARG-aware methods in human 
genomics~\cite{Mejia2025using,Grundler2025geographic,
Lehmann2025args,Terhorst2025accelerated}.
We hope that the ARG provided here will similarly
enable a range of recombination-aware analyses of SARS-CoV-2, and 
catalyse a wave of methodological developments integrating
recombination into pathogen genomics.
For instance, annotating the ARG with experimentally derived
fitness estimates for mutations can help elucidate selection
pressures on different recombinants;
analysing the genomic positions of mutations occurring on long
branches (which persist after accounting for recombination) can
illuminate the evolution of saltational variants.
We envisage the ARG as a long-term stable resource---building on the 
high-quality, freely available reassemblies of
the Viridian dataset~\cite{Hunt2024}---to facilitate the study 
of the pandemic phase of SARS-CoV-2 in an open and reproducible manner.

The other primary contribution of this work is the method used to 
infer the ARG, sc2ts. The ability to infer an ARG over millions of whole 
genomes is a step change in scale over existing ARG 
inference methods for pathogen data\cite{Didelot2010inference,%
Vaughan2017inferring,%
Muller2020bayesian,%
Ignatieva2022,%
Muller2022bayesian}.
To ensure these innovations are well-founded, we have paid significant 
attention to methodological detail, carefully validating the ARG
(and consequently the method) against known results. 
Sc2ts is currently specialised to take advantage of unique features 
of the SARS-CoV-2 dataset, but the methods may be readily adapted
to other pathogens. The LS HMM is the most important element of 
sc2ts, efficiently finding parsimonious placements for samples 
among all possible recombinant solutions. The current 
parametrisation---emphasising interpretability---is simplistic,
and could readily be extended to account
for mutation rate variation along the genome~\cite{Haddox2025}
or transition-to-transversion rate bias~\cite{Bloom2023,Simmonds2024}.
Our analysis of deletions shows that valuable phylogenetic signal
is lost when they are ignored, and an important avenue for 
future work is to explore ways of incorporating them into the HMM.
Another methodological innovation is the use of 
the tsdate method~\cite{Wohns2022,Pope2025} to estimate internal node dates
in the ARG, illustrating the reuse of methods initially developed 
for population genetic applications. While results are encouraging, 
dating of phylogenies is a complex task~\cite{Kumar2016advances}
and there is certainly room for improvement.

Further contributions of this work arise from 
the detailed study of the sc2ts ARG,
providing insights into the evolution of SARS-CoV-2
and the COVID-19 pandemic. 
First, we clarify the relationships among
recombinant Pango X lineages, finding examples of
clustering of named lineages under a single shared recombination event.
Second, we used the ARG to study breakpoint intervals 
for recombination events, adding a novel temporal dimension
and demonstrating a simple relationship with the 
size of the interval in base pairs derived from theory.
Thirdly, we demonstrate a clear relationship between
genetic diversity and the rate of detectable recombination,
with the best predictor of the number of recombination
events over time being the product of global case numbers
and lineage diversity.
Finally, and perhaps most fundamentally, through our 
conservative parameter choices and real-time handling of 
recombination events, we provide a stringent lower
bound on the amount of recombination that occurred during
the pandemic phase of SARS-CoV-2.
These results underscore the utility of the ARG 
and provide a template for future analyses in which 
% JK: Going out on a limb here, please do change if 
% we don't like it!!!
recombination is fully integrated rather than treated
as a nuisance.
